Expert-In-The-Loop Causal Discovery: Iterative Model Refinement Using Expert Knowledge

### 2024-08-07 paper/main.tex
Updates

Understanding the cause-and-effect relationships between variables is a
fundamental goal in many scientific questions as it reveals the underlying
mechanisms of observed phenomena and informs effective interventions or policy
decisions. \emph{Causal Discovery} methods aim to uncover these underlying
cause-effect relationships among random variables from observational data. The
problem of causal discovery has been explored in both the Directed Acyclic
Graphs (DAGs) and Structural Equation Models (SEMs) literatures, each adopting
a different approach. In the DAG literature, the focus has primarily been on
developing automated algorithms that can learn the causal structure from
datasets. As a result of these efforts, numerous algorithms for causal
discovery have been proposed in the DAG literature, such as constraint-based
methods like the PC algorithm \citep{Spirtes2001} and Fast Causal Inference
(FCI) \citep{Spirtes2000}), score-based methods like Hill-Climb Search and
Greedy Equivalence Search (GES) \citep{Chickering2002}, and continuous
optimization-based methods like NO TEARS \citep{Zheng2018} and DAGMA
\citep{Bello2022}. In contrast, the SEM literature has focused more towards
developing tools that can assist researchers in manually constructing or
modifying models by allowing them to integrate their expert knowledge into the
model building process. Researchers typically develop an initial model based on
domain knowledge and use various tools to guide modifications to improve the
fit of their model to data. This model construction process is known as
\emph{Specification Search}. A common method for specification search is using
\emph{modification indices}, which identify potential changes to the model
structure that improve the model fit. These potential changes are then ranked
by the ones that improve the fit the most. This ranking allows the researchers
to focus first on the suggestions that can improve the model the most and while
integrating their expertise to choose the most optimal modifications.

Even though so many automated algorithms have been proposed, the adaption of
these algorithms in applied fields remain limited, and researchers typically
prefer to construct DAGs manually based on their domain expertise
\citep{Tennant2020, Petersen2021}. We believe this is the case because of
problems with these algorithms when applied in practical settings:

\begin{enumerate}
	\item \textbf{Make Obvious Mistakes:} Although most algorithms are
		theoretically consistent asymptotically, their finite sample
		properties are not well understood. Moreover, depending on the
		selected hyperparameters, their output can vary significantly.
		These issues combined with no good method to test the
		performance of these algorithms on a given dataset, in
		real-world settings with finite samples, it can be difficult to
		trust the output of these algorithms. Consequently, researchers
		have to rely on posthoc modifications to the model based on
		their domain expertise.
	\item \textbf{Outputs Markov Equivalence Class (MEC):} Multiple
		DAGs can be faithful to a given observational dataset. Thus,
		using only observational data, algorithms can only recover the
		MECs that can contain a combination of directed and undirected
		edges. However, most methods for downstream tasks, such as
		identification or causal effect estimation, assume knowledge of
		a fully oriented DAG. As a result, researchers have to manually
		orient these edges before using them for analysis.
	\item \textbf{Difficulty in Choosing Best Algorithm and Hyperparameters:}
\end{enumerate}

@@

			\draw[->]  (age) to[bend right=20] (ed);
			\draw[->]  (sex) to (ed);
			\draw[->]  (age) to (hrpw);
			\draw[->]  (ed) to (hrpw);
			\draw[->]  (sex) to (hrpw);
			\draw[->]  (ed) to (income);
			\draw[->]  (hrpw) to (income);
			\draw[->]  (occ) to[out=300, in=0, looseness=1.4] (income.east);
			\draw[->]  (sex) to (income);
			\draw[->]  (ed) to[out=90, in=150, looseness=1.3] (occ);
			\draw[->]  (sex) to (occ);	
		\end{scope}
		\begin{scope}[xshift=-3cm]
			\draw[thick, ->, double, cyan] (2.5,0) -- (3,0);
			\node[rectangle, align=center, inner sep=1pt] at (5, 0) {
				\begin{minipage}{.2\textwidth}
					\begin{equation*}
						\begin{split}
							\textnormal{Age} &\ci \textnormal{Incm} \rvert \textnormal{Edct, Hrpw, Sex} \\
							\textnormal{Age} &\ci \textnormal{Occp} \rvert \textnormal{Edct, Sex} \\
							\textnormal{Age} &\ci \textnormal{Sex} \\
							\textnormal{Hrpw} &\ci \textnormal{Occp} \rvert \textnormal{Edct, Sex} \\
						\end{split}
					\end{equation*}
				\end{minipage}
				};
		\end{scope}
		\end{tikzpicture}
		\caption{\todo[inline]{Finish figure. Show an example of model learnt using PC algorithm $\rightarrow$ Orienting the unoriented edges $\rightarrow$  Model testing}}
		\label{fig:ci_testing}
\end{figure}
\begin{figure}
	\caption{\todo[inline]{Take a model; show the modification indices for it.}}
\end{figure}

Inspired by this ranking approach in modification indices, in this paper we
develop a method to rank the CI tests using a conditional measure of
association. Our proposed approach uses the measure of association to find
correlated variables in the data that are not explained by the current model.
Using the measure, we can then rank the pair of variables between which adding
an edge would improve the explanation the most. Based on this information, the
user can select a potential pair of variables and based on their expert
knowledge decide the direction of the edge. 
\todo[inline]{Add about the score based methods somewhere here}

>>>>

Understanding cause-and-effect relationships between variables is a fundamental
goal in many scientific inquiries, as it reveals the underlying mechanisms of
observed phenomena and informs effective interventions or policy decisions.
\emph{Causal Discovery} methods aim to uncover these relationships among random
variables from observational data. This problem has been extensively explored
in both the Directed Acyclic Graphs (DAGs) and Structural Equation Models
(SEMs) literatures, each adopting a different approach.

In the DAG literature, the focus has primarily been on developing automated
algorithms to learn causal structures from datasets. As a result of these
efforts, numerous algorithms for causal discovery have been proposed, such as
constraint-based methods like the PC algorithm \citep{Spirtes2001} and Fast
Causal Inference (FCI) \citep{Spirtes2000}), score-based methods like
Hill-Climb Search and Greedy Equivalence Search (GES) \citep{Chickering2002},
and continuous optimization-based methods like NO TEARS \citep{Zheng2018} and
DAGMA \citep{Bello2022}.

In contrast, the SEM literature has focused more on developing tools to assist
researchers in manually constructing or modifying models while allowing them to
integrate their expert knowledge into the model-building process. Researchers
typically develop an initial model based on their expertise and then use
various tools to guide modifications to improve the model's fit to data. This
model modification process is commonly known as \emph{Specification Search}
\citep{Long1983}, employing method like modification indices, and Wald-based
tests \citep{Marcoulides2018}.

Despite the many automated algorithms proposed for constructing DAGs, their
adoption in applied research remains limited, with researchers often preferring
to manually construct DAGs based on their domain expertise \citep{Tennant2020,
Petersen2021}. We believe this preference persists due to several issues with
these algorithms when applied in practical settings:

\begin{enumerate}
	\item \textbf{Obvious Mistakes:} Although most algorithms are
		theoretically consistent asymptotically, their finite sample
		properties are not well understood. Moreover, their output can
		vary significantly depending on the selected algorithm and
		their hyperparameters. Combined with the lack of robust methods
		to test these algorithms' performance on a given dataset, it
		can be difficult to trust their outputs. Consequently,
		researchers have to rely on post-hoc modifications to the model
		based on their domain expertise.
	\item \textbf{Outputs Markov Equivalence Class (MEC):} Multiple
		DAGs can be faithful to a given observational dataset. Thus,
		using only observational data, algorithms can only recover the
		MECs that can contain a combination of directed and undirected
		edges. However, most methods for downstream tasks, such as
		identification or causal effect estimation, assume knowledge of
		a fully oriented DAG. As a result, researchers have to manually
		orient these edges before using them for analysis.
\end{enumerate}

@@

			\draw[->]  (age) to[bend right=20] (ed);
			\draw[->]  (sex) to (ed);
			\draw[->]  (age) to (hrpw);
			\draw[->]  (ed) to (hrpw);
			\draw[->]  (sex) to (hrpw);
			\draw[->]  (ed) to (income);
			\draw[->]  (hrpw) to (income);
			\draw[->]  (occ) to[out=300, in=0, looseness=1.4] (income.east);
			\draw[->]  (sex) to (income);
			\draw[->]  (ed) to[out=90, in=150, looseness=1.3] (occ);
			\draw[->]  (sex) to (occ);	
		\end{scope}
		\begin{scope}[xshift=-3cm]
			\draw[thick, ->, double, cyan] (2.5,0) -- (3,0);
			\node[rectangle, align=center, inner sep=1pt] at (5, 0) {
				\begin{minipage}{.2\textwidth}
					\begin{equation*}
						\begin{split}
							\textnormal{Age} &\ci \textnormal{Incm} \rvert \textnormal{Edct, Hrpw, Sex} \\
							\textnormal{Age} &\ci \textnormal{Occp} \rvert \textnormal{Edct, Sex} \\
							\textnormal{Age} &\ci \textnormal{Sex} \\
							\textnormal{Hrpw} &\ci \textnormal{Occp} \rvert \textnormal{Edct, Sex} \\
						\end{split}
					\end{equation*}
				\end{minipage}
				};
		\end{scope}
		\end{tikzpicture}
		\caption{\todo[inline]{Finish figure. Show an example of model learnt using PC algorithm $\rightarrow$ Orienting the unoriented edges $\rightarrow$  Model testing}}
		\label{fig:ci_testing}
\end{figure}

A common method for specification search is using \emph{modification indices},
which identify potential changes to the model structure that improve the model
fit. These potential changes are then ranked by the ones that improve the fit
the most. This ranking allows the researchers to focus first on the suggestions
that can improve the model the most and while integrating their expertise to
choose the most optimal modifications. Inspired by this ranking approach in
modification indices, in this paper we develop a method to rank the CI tests
using a conditional measure of association. Our proposed approach uses the
measure of association to find correlated variables in the data that are not
explained by the current model. Using the measure, we can then rank the pair of
variables between which adding an edge would improve the explanation the most.
Based on this information, the user can select a potential pair of variables
and based on their expert knowledge decide the direction of the edge. 
\todo[inline]{Add about the score based methods somewhere here}
